Do larger watersheds respond different to forest cover change? Re-analysing a global data set.

### 2021-11-28 paper/Forest_and_Water/Forest_and_Water.Rmd
further changes to the paper

In addition, additional variables added were the latitude and longitude for the center of the watershed as an approximation of its spatial location. Using this information  reference evapotranspiration (E_{0}) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET0) Climate Databasev2 [@trabucco2018], if a value of E0 was not available from the original papers. For large watersheds, this value, similar to annual average rainfall, is only an approximation of the climate at the location.

@@

Similar to @zhang2017, the "dryness index" was calculated as:
$$\tag{1}
\begin{aligned}
D = \frac{E0}{Pa}
\end{aligned}$$

@@

<!--remove the Amazon from the analysis -->
```{r}
Zhang_all <- Zhang_all %>%
  filter(`Watershed #` > 1)
```

@@

```{r}
Zhang_all <- Zhang_all %>%
  mutate(DeltaQf_perc = ifelse(`Watershed #` == 76,157,DeltaQf_perc))
```

@@

The general model tested is

@@

Here $X_i$ are factorial variables, while $Z_i$ are continous variables. The model assumes no direct interactions and all variables are additive. The changes in forest cover contain both positive (forestation) and negative values (deforestation). In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and non-hysteretic. However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover. Therefore all the change in forest cover data is converted to positive values, and an additional column (*sign_{forest cover}*) is added that indicates whether it was a forest cover increase or decrease. 
A further assumption in the model is that all continuous variables $Z_i$ (such as annual precipitation (Pa)) can have a linear or non-linear relationship with $\Delta Q \%$. This means that a smooth function $s()$ is applied to the $Z_i$ variables.

@@

Following @zhang2017, the first step is to use a linear regression to investigate the percent change in flow as a result in the percent change forestry and modulated by the direction of the change, either an increase in forest cover, or decrease in forest cover.

@@

```{r, echo =F}
pander(round(summary(linear_model_simple)$p.table,2))
```

While the overall variance explained in this model is not high at `r round(summary(linear_model_simple)$r.sq,2)`, it clearly supports the hypothesized relationship between the change in forest cover and the change in flow. The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.5\%. However the change in flow is different for forest cover decreases compared to forest cover increases. In fact, forest cover increases decrease flow by 31\% less than a similar decrease in forest cover causes flow to increase. So roughly speaking, a 1\% forest cover increase on the average decreases flow by $(1 - 0.31)*0.5\%$, while a the percentage forest cover decrease will increase flow by 0.5\%.

It is however clear from the lack of explaining power, that there could be confounding factors, as alluded to in the methods. The obvious ones being watershed dryness and area (following @zhang2017).

@@

```{r, echo = F}
pander(round(summary(linear_model2)$p.table,2))
```

@@

As indicated, a second hypothesis relates to whether there is a strong spatial global gradient as captured by latitude and longitude. As the global map (\@ref(fig:global_map)) shows, the distribution of case study watersheds covers multiple continents and shows some distinct clustering in parts of the world. Of interest is whether the spatial clustering also indicates a difference in response to forest cover change.

@@

```{r, echo =F}
pander(round(summary(model3_includeLatLong)$p.table,2))
```

This linear model shows that there is a significant gradient in the Latitude and with annual average rainfall, with watersheds closer to the equator having lower changes in the runoff compare to watersheds further away from the equator. This suggests an influence of radiation, which will be tested next. In addition, the model suggests an influence of the annual average rainfall, with wetter watersheds having slightly lower changes in runoff. The total explaining power of the model is still low at `r round(summary(model3_includeLatLong)$r.sq,2)` suggesting further confounding factors currently not included in the model.

There is no relationship with Longitude, suggesting that the different continents do not show a trend in the East-West direction.

@@

The previous analysis suggests an influence of radiation on evapotranspiration, and most likely related to the dryness index, as also indicated in @zhang2017. Increased evapotranspiration could lead to drier watersheds, unless balanced by rainfall (such as possibly in the tropics). This model introduces the dryness index as a linear variable and drops the annual average precipitation as a variable.

@@

```{r, echo =F}
pander(round(summary(model4_dryness)$p.table,2))
```

@@

However, the result also indicates possible issues with the data, some of the Dryness values are very large (> 4) and these values have high leverage in the data. These watershed are listed in Table XX:

```{r, echo = F}
pander(Zhang_all2 %>% filter(Dryness > 4) %>% select(Latitude, Longitude,`Watershed name`))
```

@@

model5_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(Area_km2, bs="ts") + s(Dryness, bs="ts") + s(length, bs="ts") + 
                    Latitude ,  data = Zhang_all2)
summary(model5_all)
#gam.check(model5_all)
#plot(model5_all)
#pander(round(summary(model5_all)$p.table,2))
```
<!-- model 2c include smooth variables -->

@@

```{r model6_reduc}
model6_reduc <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(Dryness, bs="ts" ) + s(Latitude, bs="ts") + 
                      s(Area_km2, bs="ts") +  s(length, bs="ts") +
                    Assessment_technique +
                    Hydrological_regime, data = Zhang_all2)
summary(model6_reduc)

@@

<!-- Take out Latitude -->
```{r model7_noLat}
model7_noLat <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(Dryness, bs="ts" ) + 
                      s(Area_km2, bs="ts")  +
                    Precip_data_type +  Assessment_technique + Forest_type +
                    Hydrological_regime, data = Zhang_all2)
summary(model7_noLat)
plot(model7_noLat)
```

>>>>

In addition, additional variables added were the latitude and longitude for the center of the watershed as an approximation of its spatial location. Using this information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers. For large watersheds, this value, similar to annual average rainfall, is only an approximation of the climate at the location.

@@

Similar to @zhang2017, the "dryness index" was calculated from the reference evapotranspiration and the annual average rainfall (Pa) as:
$$\tag{1}
\begin{aligned}
D = \frac{E_{0}}{Pa}
\end{aligned}$$

@@

<!--remove the Amazon from the analysis -->
```{r, echo = F}
Zhang_all <- Zhang_all %>%
  filter(`Watershed #` > 1)
```

@@

```{r,echo = F}
Zhang_all <- Zhang_all %>%
  mutate(DeltaQf_perc = ifelse(`Watershed #` == 76,157,DeltaQf_perc))
```

@@

The general model tested is:

@@

Here $X_i$ are factorial variables, while $Z_i$ are continous variables. The model assumes no direct interactions and all variables are additive. The changes in forest cover contain both positive (forestation) and negative values (deforestation). In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and non-hysteretic. However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover. Therefore all the change in forest cover data is converted to positive values, and an additional column (*sign_{forest cover}*) is added that indicates whether it was a forest cover increase or decrease. 
A further assumption in the model is that all continuous variables $Z_i$ (such as annual precipitation (Pa)) can have a linear or non-linear relationship with $\Delta Q \%$. This means that a smooth function $s()$ is applied to the $Z_i$ variables. For the smoothing function we applied thin plate regression splines with an additional shrinkage penalty which means the terms can be shrunk to 0 if not significant [@wood2006].

@@

Following @zhang2017, the first step is to use a linear regression to investigate the percent change in flow as a result in the percent change forestry and modulated by the direction of the change, either an increase in forest cover, or decrease in forest cover:

$$\tag{3}
\begin{aligned}
\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
\end{aligned}$$

@@

```{r, echo =F}
pander(round(summary(linear_model_simple)$p.table,2), caption="Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change")
```

While the overall variance explained in this model is not high with an adjusted *r^2^* of `r round(summary(linear_model_simple)$r.sq,2)`, it clearly supports the hypothesized relationship between the change in forest cover and the change in flow. The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.5\%. However the change in flow is different for forest cover decreases compared to forest cover increases. In fact, forest cover increases decrease flow by 31\% less than a similar decrease in forest cover causes flow to increase. So roughly speaking, a 1\% forest cover increase on the average decreases flow by $(1 - 0.31)*0.5\%$, while a the percentage forest cover decrease will increase flow by 0.5\%.

It is however it is clear from the lack of explaining power, that there could be confounding factors, as alluded to in the methods. The obvious ones being watershed dryness and area (following @zhang2017):

$$\tag{4}
\begin{aligned}
\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + Area_{km^2} + \varepsilon
\end{aligned}$$
Where $Pa_mm$ is the annual average rainfall in mm.

@@

```{r, echo = F}
pander(round(summary(linear_model2)$p.table,2), caption="Summary of the second model, taking into account the annual rainfall and the area of the watershed")
```

@@

As indicated, a further hypothesis relates to whether there is a strong spatial global gradient as captured by latitude and longitude. As the global map (\@ref(fig:global_map)) shows, the distribution of case study watersheds covers multiple continents and shows some distinct clustering in parts of the world. Of interest is whether the spatial clustering also indicates a difference in response to forest cover change:

$$\tag{5}
\begin{aligned}
\Delta \% Q \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + Area_{km^2} + Latitude + Longitude + \varepsilon
\end{aligned}$$

@@

```{r, echo =F}
pander(round(summary(model3_includeLatLong)$p.table,2), caption= "Results of the model including Latitude and Longitude")
```

This linear model shows that there is a significant gradient in the Latitude and with annual average rainfall, with watersheds closer to the equator having lower changes in the runoff compare to watersheds further away from the equator. This suggests an influence of radiation, which will be tested next. In addition, the model suggests an influence of the annual average rainfall, with wetter watersheds having slightly lower changes in runoff. The total explaining power of the model is still low with an adjusted *r^2^* of `r round(summary(model3_includeLatLong)$r.sq,2)` suggesting further factors that are currently not included in the model.

There is no relationship with Longitude, suggesting that the watersheds across different continents do not show an East-West direction trend.

@@

Latitude might be indicating an influence of radiation on evapotranspiration, and most likely related to the dryness index, as also indicated in @zhang2017. Increased evapotranspiration could lead to drier watersheds, unless balanced by rainfall (such as possibly in the tropics). The North-South trend in dryness would be related to Latitude. This model introduces the dryness index as a linear variable and drops the annual average precipitation as a variable, as dryness is calculated from.

@@

```{r, echo =F}
pander(round(summary(model4_dryness)$p.table,2), caption = "Results of the model replacing the annual precipitation with the dryness index")
```

@@

However, the result also indicates possible issues with the data, some of the Dryness values are very large (> 4) and these values have high leverage in the data. These watersheds are listed in Table 5:

```{r, echo = F}
pander(Zhang_all2 %>% filter(Dryness > 4) %>% select(Latitude, Longitude,`Watershed name`), caption = "Watersheds for which the dryness inddex > 4")
```

### Are some of the variables possibly non-linear?

The work by @filoso2017 and earlier by @jackson2005 has indicated that the length of the study might influence the response. This links to the idea from @kuczera1987 that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also @jackon2005). In addition to adding the length as a variable, all continuous variables (Dryness, Area, length and Latitude) were considered non-linear in this model and as indicated a shrinkage smoothing spline [@wood2006] was applied to these variables.

@@

model5_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(Area_km2, bs="ts") + s(Dryness, bs="ts") + 
                    s(length, bs="ts") + s(Latitude, bs = "ts") ,
                  data = Zhang_all2)
summary(model5_all)
#gam.check(model5_all)
#plot(model5_all)
pander(round(summary(model5_all)$p.table,2))
pander(round(summary(model5_all)$s.table,2))
```
<!-- model 2c include smooth variables -->

The overall explaining power of the model increases with an adjusted *r^2^* of `r round(summary(model5_all)$r.sq,2)`, but little changes in the significance of the variables.

A final full model

@@

```{r model6_reduc, eval = F}
model6_reduc <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(Dryness, bs="ts" ) + s(Latitude, bs="ts") + 
                      s(Area_km2, bs="ts") +  s(length, bs="ts") +
                    Assessment_technique +
                    Hydrological_regime, data = Zhang_all2)
summary(model6_reduc)

@@

<!-- Take out Latitude -->
```{r model7_noLat}
model7_noLat <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(Dryness, bs="ts" ) + 
                      s(Area_km2, bs="ts")  + s(length, bs = "ts") +
                    Precip_data_type +  Assessment_technique + Forest_type +
                    Hydrological_regime, data = Zhang_all2)
summary(model7_noLat)
plot(model7_noLat)
```

```{r model7_noLat_intLength}
model7_noLat_intlength <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(Dryness, bs="tp" ) + 
                      s(Area_km2, bs="tp")  + s(length, bs = "tp") +
                    Precip_data_type +  Assessment_technique + Forest_type +
                    Hydrological_regime, 
                    data = Zhang_all2)
summary(model7_noLat_intlength)
plot(model7_noLat_intlength)
```

The flexible nature of the splines and the shrinkage applied to the splines means that the length variable can capture variation in the data, but it is unclear what exactly is captured. The shape of the resulting conditional response does not reflect the reponses highlighted in @filoso2017 and @jackson2005. Reducing the flexibility of the splines, or fitting a linear term results in "length" not being significant, which suggests that this variable is not helpful in explaining the variation in the data.
